### 2008-05-30 trunk/Papers/rmrs-08/rmrs.tex
very prelim version

Define syntax of RMRS; model theory of RMRS; solved forms.  Explain
the role of the RMRS models (= trees) in the semantic construction
pipeline, in particular with respect to decoding trees into semantic
representations of the base language formalism.  Present examples in
MRS syntax first, then in our RMRS language.  Finally, compare our
version of RMRS against MRS and dominance constraints.

>>>>

% Define syntax of RMRS; model theory of RMRS; solved forms.  Explain
% the role of the RMRS models (= trees) in the semantic construction
% pipeline, in particular with respect to decoding trees into semantic
% representations of the base language formalism.  Present examples in
% MRS syntax first, then in our RMRS language.  Finally, compare our
% version of RMRS against MRS and dominance constraints.

\subsection{RMRS Syntax}

To formalise the basic ideas from Section~\ref{sec:motivation}, 
we start by defining the syntax of the {\sc rmrs} language.
We adopt a
syntax in the style of {\sc clls} (ref), in that labels, anchors and
holes are all represented as {\em node variables}
$\nvar=\{X,Y,X_1,\ldots\}$---the distinction among them comes
from their position in the formulae of an {\sc rmrs} description (see
Definition~\ref{defn:rmrs-syntax}). 
The {\sc rmrs} vocabulary also includes
base variables $\ovar$, consisting of
individual variables $\{x,x_1,y,\ldots\}$ and event variables
$\{e_1,\ldots\}$ (one can also include supersorts to
these, but we forego doing so here).  Finally, the vocabulary includes
0-ary predicate symbols $\pred=\{P,Q,P_1,\ldots\}$.

\begin{definition}\label{defn:rmrs-syntax}
An {\sc rmrs} is an underspecified description $\varphi$ consisting of a
finite set of atoms 
of the forms in (\ref{eg:atomic-forms}), such that $\varphi$
satisfies the
following three constraints:
\begin{enumerate}
\item   If $\ep{X}{Y}{P}, \ep{Z}{Y}{Q}\in \varphi$, then $X=Z$ and $P=Q$
  (i.e., each predication in $\varphi$ has a unique anchor);
\item   If $\ARG_S(X,v)\in \varphi$, then $X:Y:P\not\in \varphi$ and
  $ARG_{S'}(Y,X)\not\in \varphi$ for any $Y\in \nvar$ and
  $P\in \pred$ (i.e., the first argument to an ARG-relation must be an
  anchor).
\item   If $\ep{X}{Y}{P}\in \varphi$, then $\ep{Y}{Z}{Q},
  \ep{Z}{X}{Q}\not\in \varphi$ for all 
  $Z\in \nvar$ or $Q\in \pred$ (i.e., an anchor cannot be a label, or
  {\em vice versa}).
\end{enumerate}
\begin{examples}
\item   \label{eg:atomic-forms}
$\begin{array}{rrll}
\varphi &::= &\ep{X}{Y}{P} \\ %& \mbox{labelled elementary predication} \\
&| &\ARG_S(X,v) \\ % & \mbox{ARG relation for base variables} \\
&| &\ARG_S(X,Y) \\ % & \mbox{ARG relation for node variables} \\
&| &X \qeq Y \\ % & \mbox{QEQ} \\
&| &v_1 = v_2 \;|\; v_1 \neq v_2 \\ % & \mbox{base variable
                                % (in)equality} \\
&| &X = Y \;|\; X \neq Y \\ % & \mbox{node variable (in)equality} \\
&| &P \spec Q % & \mbox{SPEC relation between predicate symbols}
\end{array}
$
\end{examples}
\end{definition}

The subscript $S$ in the ARG relations stands for a subset of $\N$ which
is either finite or $\N$ 
itself.  The original {\sc rmrs} syntax $\ARG_i$ is 
an abbreviation of $\ARG_{\{i\}}$ and $\ARG_n$ is an abbreviation of
$\ARG_\N$.  Furthermore, $l:a:P(v)$ in the {\sc rmrs}s from
Section~\ref{sec:motivation} and earlier papers (e.g.,
\cite{copestake:2003}) 
is a notational variant of $\{X:Y:P, ARG_{\{0\}}(v)\}$ in
Definition~\ref{defn:rmrs-syntax}.

\subsection{Model Theory}

The model theory formalises the relationship between an {\sc rmrs} and
the fully specific, alternative logical forms that it describes, as
expressed in the base language.
As suggested in Section~\ref{sec:motivation},  each model $\tau$
corresponds to a unique formula of the base language, conceived as a
tree (see Figure~\ref{fig:1}).  {\em Satisfaction} must then give rise
to a notion of {\em 
  truth} and {\em entailment} that respectively correspond to an {\sc
  rmrs} $\varphi$ being a partial description of $\tau$, and one {\sc
  rmrs} $\varphi$ being more specific than another $\varphi'$ (in the sense that
$\varphi$ imposes more
constraints than $\varphi'$ on the possible fully-specific logical forms).

Since each model corresponds to a (unique) base language formula, defining
models depends on the base-language's vocabulary $\Sigma$.  More
formally, $\Sigma$ is a sorted signature of
base-language constructors $f$ of sorts $s_1\times\ldots
s_n\rightarrow s$,
where $s,s_1,\ldots,s_n\in {\cal S}$, ${\cal S}$ being the sort
hierarchy for $\Sigma$.  We call $n\geq 0$ the {\em arity} of $f$, and
we write $\Sigma_i$ and $\Sigma_{\geq i}$ for the constructors in $\Sigma$
with arity exactly $i$ and at least $i$ respectively.

The models of {\sc rmrs} are then defined to be finite constructor
trees:
\begin{definition}\label{defn:models}
A {\em finite constructor tree} $\tau$ is
a function $\tau:D
\rightarrow \Sigma$ such that $D$ is a \emph{tree domain} (i.e., a subset of
$\N^*$ which is closed under prefix and left sibling) and the number
of children of each node $u \in D$ is equal to the arity of $\tau(u)$.
We define the \emph{sort} $s(u)$ of each node $u$ in $\tau$ bottom-up
by saying that if $u$ has children $u_1,\ldots,u_n$ and $s(\tau(u)) =
s_1 \times \ldots \times s_n \rightarrow s$, then $s(u) = s$ iff
$s(u_i) \leq s_i$ for all $i$.  $\tau$ is \emph{well-sorted} if
all nodes can be assigned sorts in this way.  In what follows, we only consider
well-sorted finite constructor trees.
\end{definition}

The same tree $\tau$ can
represent formulas from different-style base languages (e.g.,
extensional and modal languages, {\sc drt}, etc) by 
leaving the encoding flexible.  But for reasons 
of space, we don't
explore this here; for simplicity, we
assume a sort hierarchy ${\cal S}$ for $\Sigma$ that is
closely related to the sorts of {\sc rmrs} descriptions
themselves.  So ${\cal S}$ contains the sorts
$h$ (for subformulas), $x$ (for individual variables) and $e$ (for event
variables).
We also
assume that $\Sigma$ contains conjunction constructors
$\wedge_i: h\times\ldots h\rightarrow h$ of arity $i$ for every
$i\in \N$.  We write $D(\tau)$ for the tree domain of a
constructor tree $\tau$, and further define the following relations
between nodes in a finite constructor tree:

\begin{definition}\label{defn:dominance}
$u \qeqdom v$ iff $v = u \cdot 2^k$ for some $k \geq 0$ and all
  symbols $\tau(u), \tau(u \cdot 2), \ldots, \tau(u \cdot 2^{k-1})$
  are quantifiers.

$u \wedgedom v$ iff $u \leq v$ (i.e.\ it is a prefix), and all
  symbols on the path from $u$ to $v$ (not including $v$) are one of
  the $\wedge_i \in \Sigma$.
\end{definition}

The {\em satisfaction} relation between an {\sc rmrs} description $\varphi$
and a finite constructor tree $\tau$ is defined in terms of several
assignment functions.  The need for a 
node variable
assignment function
$\alpha:
\nvar
\rightarrow D(\tau)$ is clear; the
constraints on scope defined by the {\sc rmrs} must be resolvable to
$\tau$'s specific
specific scope information, and for any given $\tau$ there may be more
than one way to do this.
Secondly, to ensure that, for
instance, the predications corresponding to {\em every} and {\em cat}
in the {\sc rmrs}
(\ref{eg:cat-pos}b)  partially describe fully-specific
logical forms that accurately record the semantic dependency between
these lexemes in (\ref{eg:cat}a) as well as partially describing
fully-specific logical forms where they are not semantically
dependent (e.g., as required when the same predications are produced
from the {\sc pos} tagged sentence {\em every dog chased some cat}),
the relationship of satisfaction 
should ensure that the variables from $\ovar$ in an
{\sc rmrs} description can
map in a many-to-one way to variables in $D(\tau)$.  So
satisfaction must also be defined in
terms of 
a (0-ary) variable assignment function
$g:\ovar \rightarrow \Sigma_0$ from {\sc rmrs} variables to 
base-language variables that respects the sorts, i.e.\
$s(g(v)) \leq s(v)$ for all $v \in \ovar$.  Finally, satisfaction must
be defined in terms of a relation $\sigma \subseteq
\pred \times \Sigma_{\geq 1}$ that maps each
{\sc rmrs} predicate symbol to a set of constructors from $\Sigma$.

\begin{definition}\label{defn:satisfaction}
Satisfaction of atoms is defined as follows:
$$
\begin{array}{l}
  \tau, \alpha, g, \sigma \models  \ep{X}{Y}{P}
\mbox{ iff}\\
\qquad \tau(\alpha(Y)) \in \sigma(P) \mbox{ and } \alpha(X) \wedgedom
  \alpha(Y) \\
  \tau, \alpha, g, \sigma \models \ARG_S(X,a)
\mbox{ iff exists  $i \in S$ s.t.}\\
\qquad  \alpha(X) \cdot i \in D(\tau) \mbox{ and }
  \tau(\alpha(X) \cdot 
  i) = g(a) \\
  \tau, \alpha, g, \sigma \models \ARG_S(X,Y)
\mbox{ iff  exists $i \in S$ s.t.}\\
\qquad \alpha(X) \cdot i \in D(\tau), \alpha(X) \cdot
  i = \alpha(Y) \mbox{ and }\\
\qquad
s(\alpha(X) \cdot i) = h \\
  \tau, \alpha, g, \sigma \models X \qeq Y
\mbox{ iff } \alpha(X) \qeqdom \alpha(Y) \\
  \tau, \alpha, g, \sigma \models X ={/}\neq Y
\mbox{ iff } \\
\qquad \alpha(X) ={/}\neq \alpha(Y) \\
  \tau, \alpha, g, \sigma \models v_1 ={/}\neq v_2
\mbox{ iff}\\
\qquad g(v_1) ={/}\neq g(v_2) \\
  \tau, \alpha, g, \sigma \models P \spec Q
\mbox{iff $\sigma(P) \subseteq \sigma(Q)$}
\end{array}
$$
A 4-tuple $\tau,\alpha,g,\sigma$ satisfies an underspecified
representation $\varphi$ (written $\tau,\alpha,g,\sigma \models \varphi$)
iff it satisfies all of its
elements.
\end{definition}
Satisfaction encapsulates information about
semantic scope ambiguities, missing
information about 
semantic dependencies and/or missing information about lexical
subcategorisation and lexical senses.   For $j=\{1,2\}$, suppose that
$\tau_j,\alpha_j,g_j,\sigma\models \varphi$ and let $\vocab(\varphi)$
be the vocabulary that features in $\varphi$.  Then 
$\varphi$ exhibits a semantic scope
ambiguity if there is a $Y,Y'\in \vocab(\varphi)$ such that $\alpha_1(Y) \qeqdom
\alpha_1(Y')$ and $\alpha_2(Y')\qeqdom \alpha_2(Y)$.   It exhibits
missing information about semantic dependencies if $\exists v,v'\in
\ovar$ in $\vocab(\varphi)$ such that $g_1(v)=g_1(v')$ and
$g_2(v)\neq g_2(v')$.  It 
exhibits missing lexical subcategorisation information if $\exists
Y\in \vocab(\varphi)$ such that $\tau_1(\alpha_1(Y))$ is a constructor of a
different type from $\tau_2(\alpha_2(Y))$.  And it exhibits missing
lexical sense information if $\tau_1(\alpha_1(Y))$ and
$\tau_2(\alpha_2(Y))$ are different base-language constructors, but of
the same type.

Truth, validity and entailment can now be defined in terms of
satisfiability 
in the usual way:
\begin{definition}\label{defn:entailment}
Let $\tau$ be a model, $\alpha$ a node variable assignment function,
$g$ a 0-ary variable assignment function, $\sigma$ a mapping from
$\pred$ to the powerset of $\Sigma$, and $\varphi,\varphi'$ {\sc rmrs}
descriptions.  Then:
\begin{description}
\item   [truth:] $\tau\models \varphi$ iff $\exists \alpha,g,\sigma$  such
  that $\tau,\alpha,g,\sigma\models \varphi$
\item   [validity:] $\models \varphi$ iff $\forall \tau$, $\tau\models \varphi$.
\item   [entailment:] $\varphi\models \varphi'$ iff $\forall \tau$, if
  $\tau\models \varphi$ then $\tau\models \varphi'$.
\end{description}
\end{definition}
% NOTE: This definition of entailment isn't right, because it allows
% the assignment $\sigma$ from predicates to constructors to vary
% when testing the satisfiability of $\varphi$ vs. $\varphi'$ with
% respect to $\tau$.  I think that to get the notion of entailment we
% want, $\sigma$ should actually be part of the {\sc rmrs} *model*.
% In other words, an {\sc rmrs} model $M$ is a tuple $\langle \tau,
% \sigma\rangle$, where $\tau$ is a finte constructor tree and
% $\sigma$ is a mapping from $\Pred$ to the power set of $\Sigma$.
% Then truth and entailment are defined correctly.  I believe our
% proofs of things like Theorem~\ref{thm:big-one} actually assumes
% that $\sigma$ is constant when interpreting $\varphi$ vs.\
% $\varphi'$, no?

\subsection{Solved Forms}

Following the framework of dominance constraints, one can define a fragment
of {\sc rmrs} descriptions, known as {\em solved forms}, which are
guaranteed to be satisfiable.  For dominance constraints, solved forms
are sets of constraints whose node variables form a forrest under the
dominance relation $\dom$.  Solved forms for {\sc rmrs} must impose
additional constraints, so as to ensure that all the (underspecified)
ARG relations are satisfiable.  Indeed, we help to ensure this by
substituting assuming that {\sc rmrs} solved forms have any variable
equalities substituted into the ARG relations themselves.

\begin{definition}\label{defn:solved-forms}
  An RMRS $\varphi$ is \emph{in solved form} iff:
  \begin{enumerate}
  \item every variable in $\varphi$ is either a hole, a label or an
    anchor (but not two of 
    these);
%% NOTE: The more constrained definition of an {\sc rmrs} description
%% that we have now makes the set of labels and anchors mutually
%% exclusive.  It does not make labels and holes mutually exclusive;
%% nor anchors and holes mutually exclusive
%% however.  
  \item $\varphi$ doesn't contain equality, inequality, and SPEC
    ($\sqsubseteq$) atoms;
%   \item no downward-looking label occurs on the left-hand side of two
%    different EPs;
  \item if $\ARG_S(Y,i)$ is in $\varphi$, then $|S| = 1$;
  \item for any label $Y$ and index set $S$, there
    are no two atoms $\ARG_S(Y,i)$ and $\ARG_S(Y,i')$ in $\varphi$;
  \item no label occurs on the right-hand side of two
    different $\qeqdom$ atoms.
  \end{enumerate}
\end{definition}

\begin{claim}
  Every RMRS in solved form is satisfiable.
\end{claim}
\begin{proof}
by constructing a model $\tau$ that
contains for each predicate $P\in \varphi$ a member of $\sigma(P)$ with
maximum arity, and that satisfies the ARGs and dominances in $\varphi$.
\end{proof}

One can now define the solved forms of an {\sc rmrs}
description:
\begin{definition}  \label{defn:solved-form-of}
  The \emph{syntactic dominance relation} in an {\sc  rmrs} $\varphi$ is the
  reflexive, transitive closure of the binary relation $$
\begin{array}{r c l}
\{(X,Y) & \mid &  \mbox{$\varphi$ contains $X \qeqdom Y$ or}\\
&& \mbox{$ARG_S(X,Y)$ for some
    $S$}\}
\end{array}
$$  
We write $D(\varphi)$ for the syntactic dominance
  relation of $\varphi$.

  An RMRS $\varphi'$ is \emph{a solved form of} the RMRS $\varphi$
  iff $\varphi'$ is in solved form and there is a substitution $s$
  that maps the  node and base variables of $\varphi$ to the node and
  base variables of $\varphi'$ such that
  \begin{enumerate}
  \item $\varphi'$ contains the EP $\ep{X'}{Y'}{P}$ iff there are
    variables $X,Y$ such that $\ep{X}{Y}{P}$ is in $\varphi$, $X' =
    s(X)$, and $Y' = s(Y)$;
  \item for every atom $\ARG_S(X,i)$ in $\varphi$, there is exactly
    one atom $\ARG_{S'}(X',i')$ in $\varphi$ with $X' = s(X)$, $i' =
    s(i)$, and $S' \subseteq S$;
  \item $D(\varphi') \supseteq s(D(\varphi))$.
  \end{enumerate}
\end{definition}

\begin{claim}
  For every model $\tau$ of some {\sc rmrs} $\varphi$, there is a solved form
  $\varphi'$ of $\varphi$ such that $\tau$ is also a model of $\varphi'$.
\end{claim}
\begin{proof}
  will be by constructing a solved form that respects the equalities,
  ARGs, and dominances in $\tau$.
\end{proof}

\begin{prop}
  Every {\sc rmrs} $\varphi$ has only a finite number of solved forms, up to
  renaming of variables.
\end{prop}
\begin{proof}
  Up to renaming of variables, there is only a finite number of
  substitutions on the node and base variables of $\varphi$.  Let $s$
  be such a substitution.  This fixes the set of EPs of any solved
  form of $\varphi$ that is based on $s$ uniquely.  There is only a
  finite set of choices for the subsets $S'$ in condition 2 of
  Def.~\ref{defn:solved-form-of}, and there is only a finite set of
  choices of new $\qeqdom$ atoms that satisfy condition 3.  Therefore,
  the set of solved forms of $\varphi$ is finite.
\end{proof}
